---
title: "Hackathon November 2021 - Group 2"
output: 
  html_document:
    css: ["style.css", "colours.css"]
---

```{r, echo=FALSE}
options(scipen = 999)

library(openxlsx)
library(data.table)
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)
library(rmarkdown)
```

### Introduction

We have been given the task of exploring time series analysis as part of a hackathon. We have been given access to 3 datasets:

1. Mortality data
2. [Data on COVID-19 (coronavirus) by Our World in Data](https://github.com/owid/covid-19-data/tree/master/public/data)
3.

### Exploratory data analysis

###### 1 Mortalilty data

Read in data and format as time series.

```{r}
# Read in the correct sheet as a data.table
mortality_df <- 
  read.xlsx(
    xlsxFile = "Data/MortalityDataEngandWales_2011_2021.xlsx", 
    sheet = "TotalDeaths2011to2020"
  ) %>% 
  as.data.table()

# Name the table
setnames(mortality_df, 1:2, c("WE","Deaths"))

# Convert WE to a date
mortality_df[, WE := as.Date(WE, origin = as.Date("1899-12-30"))]

# Create the time series
mortality_ts <- ts(
  data = mortality_df$Deaths, 
  freq = 365.25/7, # https://stackoverflow.com/questions/22188660/r-time-series-modeling-on-weekly-data-using-ts-object
  start = decimal_date(ymd("2011-01-07"))
) 

```

The first confirmed cases of coronavirus in the UK were on January 29 [2020], when two Chinese nationals fell ill at the Staycity Aparthotel in York.

Plot the time series.

```{r, out.width="100%"}
p <- mortality_df %>%
  ggplot(aes(WE, Deaths)) + 
  geom_line() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) x/1e3) +
  geom_vline(xintercept = as.numeric(ymd("2020-01-29")), color="red", lty=2) +
  ylab("Deaths (thousands)") +
  theme_minimal() +
  theme(axis.title.x = element_blank())

ggplotly(p)
```

```{r}

#2/3 of observations of the covid-free period
mortality_train_ts <- round(mortality_df[WE < ymd("2020-01-29"), .N] * 2/3)

mortality_train_ts <- mortality_ts[1:mortality_train_ts]

```

###### 2. Data on COVID-19 (coronavirus) by Our World in Data

This data can be read in directly from GitHub using `{readr}`. The `col_types` have to be specified up front otherwise it doesn't load correctly.

```{r}
# Read in COVID data from GitHub with correct types
covid_df <- readr::read_csv(
  file = "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv",
  col_types = "cccDdddddddddddddddddddddddddddddcddddddddddddddddddddddddddddddddd"
)

# Inspect it
paged_table(covid_df)
```

There are many columns and groups but we are going to focus on United Kingdom with a subset of columns.

```{r}
# Filter and subset columns
uk_covid_df <- covid_df %>% 
  filter(location == "United Kingdom") %>%
  # Rename some columns so they are either {col} or {col}_cumulative
  select(
    date, 
    cases = new_cases,
    cases_cumulative = total_cases,
    deaths = new_deaths, 
    deaths_cumulative = total_deaths, 
    #excess_mortality, # this is the % difference 
    excess_mortality_cumulative = excess_mortality_cumulative_absolute)

# Inspect it
paged_table(uk_covid_df)
```

We can see that the `excess_mortality_cumulative` is only updated weekly. So we forward `fill`.

```{r}
# Replace missing data from top to bottom
uk_covid_df <- uk_covid_df  %>%
  fill(excess_mortality_cumulative)

# Check it's worked
paged_table(uk_covid_df)
```

We also reverse engineer an `excess_mortality` column.

```{r}
# Reverse engineer the cumulative sum 
uk_covid_df <- uk_covid_df %>%
  # Replace all NAs with 0
  replace(is.na(.), 0) %>%
  mutate(
    excess_mortalilty = 
      excess_mortality_cumulative - lag(excess_mortality_cumulative)
  ) %>%
  # Catch the first NA that has become 0 because of the lag calc
  replace(is.na(.), 0) %>%
  dplyr::relocate(excess_mortalilty, .before = excess_mortality_cumulative)

# Inspect
paged_table(uk_covid_df)
```

Check the date range.

```{r,collapse=TRUE}
# Print off a summary of the dates
print(paste("Start date of cases:",uk_covid_df %>% summarise(min(date)) %>% pull()))
print(paste("Start date of deaths:", uk_covid_df %>% filter(deaths_cumulative > 0) %>% summarise(min(date)) %>% pull()))
print(paste("End date of cases:",uk_covid_df %>% summarise(max(date)) %>% pull()))
print(paste("End date of deaths:", uk_covid_df %>% filter(deaths_cumulative > 0) %>% summarise(max(date)) %>% pull()))
```

Quickly plot some of the cols (`{col}` on the right and `{col}_cumulative` on the right).

```{r, out.width="100%"}
# Chart the cols (not sure how you can have negative deaths)
p <- uk_covid_df %>% 
  pivot_longer(cols = -date, names_to = "measure", values_to = "count") %>%
  mutate(measure = as.factor(measure)) %>%
  ggplot(aes(x = date, y = count)) +
  geom_line() +
  facet_wrap(facets = ~measure, ncol = 2, scales = "free_y", shrink = FALSE) + 
  theme_minimal()

ggplotly(p)
```
